Unbalanced Mallows Models for Optimizing Expensive Black-Box Permutation Problems

### 2021-02-04 paper/main.tex
conclusions

In this paper, we have introduced UMM, a population-based probabilistic
algorithm based on an unbalanced Mallows model. The algorithm is designed for
expensive black-box combinatorial problems on permutation landscapes when the
budget of fitness evaluations is severely limited (here, up to $100$, $200$ and
$400$ evaluations).  Our experiments presented here show that UMM is not far
away from, and sometimes  is able to match, the results of the state-of-the-art
Bayesian optimizer for expensive black-box optimization.
%if the computation time of the optimizers is not a concern.

In addition, UMM is specially well-suited for budget-limited scenarios that are
still time-sensitive, e.g., when the overhead incurred by the optimizer should
be no more than a few minutes added to each fitness evaluation. By comparison,
CEGO may require around one day to perform 400 fitness evaluations, with the time
increasing non-linearly with permutation size and larger budgets.  Hence, UMM
is a computationally feasible alternative for relatively large problem sizes.

In this paper, we discuss the importance of the representation of permutations in applied mathematics problems. In the context considered in this paper, permutations can bee seen as rankings or orderings of items. In other problems they could be matchings or cycles. 
While certain functions are
this discussion is usually omitted in the literature.
We discuss the different concepts and show, experimentally, the consequences of misunderstanding both.

Although we were inspired by previous work on ACO for expensive black-box
combinatorial problems~\citep{PerLopStu2015si}, UMM is, to the best of our
knowledge, the first EDA specifically designed for such problems that is able
to match (and sometimes even improve) the state-of-the-art
optimizer. Moreover, it is overwhelmingly faster than the alternatives for
large problem sizes.  The results presented here show that, despite the
intrinsic difficulty of the expensive black-box combinatorial scenario, there
are still significant advances yet to be made. Thus, we hope that this paper
will motivate further research.

In this paper, we focus on the UMM under the \ken distance. Therefore, we consider appropriate problems for it, such as the Linear Ordering Problem and the Permutation Flow-shop Scheduling Problem, where the pairwise ordering among the items is relevant (as opposed to the actual exact position of each item in the permutation)

However, we are certain it is
possible to extend UMM to other distance metrics, which will allow us to
dynamically select among various distance metrics for an unknown black-box permutation landscape.
Also, as shown by our
experiments, the behavior of the parameters ($r_1$ and $r_2$) of UMM is not
always obvious, thus a more detailed analysis would be needed to provide either
generally good static values or an online adaptation approach. Finally, we have
used here  different combinatorial problems (LOP, PFSP) as
black-box benchmarks. However, an even more diverse range of problems would be
needed to understand the behavior of UMM on real-world black-box combinatorial
landscapes.

>>>>

In this paper, we have introduced UMM, a population-based probabilistic
algorithm based on an unbalanced Mallows model. The algorithm is designed for
expensive black-box combinatorial problems on permutation landscapes when the budget
of fitness evaluations is severely limited (here, up to $100$, $200$ and $400$ evaluations).
Although we were inspired by previous work on ACO for expensive black-box
combinatorial problems~\citep{PerLopStu2015si}, UMM is, to the best of our
knowledge, the first EDA specifically designed for such problems.

We also discuss the importance of the representation of permutations in applied mathematics problems. Permutations can bee seen as rankings or orderings of items and this difference is often neglected. Based on this differentiation, we propose an improvement for CEGO over the original version. The results of UMM clearly improve over the original CEGO and they are competitive with the improved version of CEGO proposed here.

In addition, UMM is specially well-suited for budget-limited scenarios that are still time-sensitive, e.g., when the overhead incurred by the optimizer should
be no more than a few second per fitness evaluation (no more than one hour for $400$ evaluations), due to its quasi-linear time complexity.
By comparison, CEGO may require several minutes per fitness evaluation (around 22 hours for $400$ evaluations), with the time increasing non-linearly with permutation size and larger budgets.  Hence, UMM is a computationally feasible alternative for relatively large problem sizes.

The current version of UMM uses the \ken distance and, therefore, it is appropriate for problems where the pairwise ordering among the items is relevant (as opposed to the actual exact position of each item in the permutation), such as the Linear Ordering Problem and the Permutation Flow-shop Scheduling.
However, we are certain that it is
possible to extend UMM to other distance metrics such as Hamming, which will allow us to
 dynamically select among various distance metrics for an unknown black-box permutation landscape.
%Also, as shown by our
%experiments, the behavior of the parameters ($r_1$ and $r_2$) of UMM is not always obvious, thus
In addition, a more detailed analysis would be needed of the parameters ($r_1$ and $r_2$) of UMM to provide either
generally good static values or an online adaptation approach.
Finally, an even more diverse range of problems would be
needed to understand the behavior of UMM on real-world black-box combinatorial
landscapes.

The results presented here show that, despite the intrinsic difficulty of the expensive black-box combinatorial scenario, there are still significant advances yet to be made. Thus, we hope that this paper will motivate further research.
